---
title: "Installing DoubleML"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing DoubleML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = FALSE)
```


## Installation 

Please note that the current version of the `DoubleML` package for R is still under development and that the code might be subject to changes. 

### Installation from CRAN

For installation from CRAN type
```{r, eval = FALSE}
install.packages("DoubleML")
```

Load the package after completed installation. 

```{r, message=FALSE, warning=FALSE}
library(DoubleML)
```

### Installation from GitHub

The `DoubleML` package for R can be downloaded using (previous installation of the [`remotes` package](https://remotes.r-lib.org/index.html) is required).

```{r, eval = FALSE}
remotes::install_github("DoubleML/doubleml-for-r")
```

Load the package after completed installation. 

```{r, message=FALSE, warning=FALSE}
library(DoubleML)
```

The python package `DoubleML` is available via the github repository. For more information, please visit our [user guide](https://docs.doubleml.org/). 
